The Impact of Malaria Episodes and Treatment Regimens on Adverse Pregnancy Outcomes in Ugandan Women

### 2025-03-11 products/manuscript/Manuscript.qmd
Project Update

---
title: "The Impact of Malaria Episodes and Treatment Regimens on Adverse Pregnancy Outcomes in Ugandan Women"
author: "Asmith Joseph"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
    prefer-html: true
    reference-location: margin  # Enable reference links
knitr:
  opts_chunk:
    screenshot.force: TRUE
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

@@

>>>>

---
knitr:
  opts_chunk:
    echo: false
    message: false
    warning: false

title: "The Impact of Malaria Episodes and Treatment Regimens on Adverse Pregnancy Outcomes in Ugandan Women"
author: "Asmith Joseph"
format:
  docx:
    toc: false
    number-sections: true
    reference-location: margin
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load Required Libraries
# Essential libraries for data handling, visualization, and analysis

library(here)             # Manage file paths
library(tidyverse)        # Core data wrangling: includes dplyr, ggplot2, readr, forcats, lubridate
library(janitor)          # Cleaning column names
library(skimr)            # Quick data summary
library(gtsummary)        # Creating summary tables
library(gt)               # Formatting tables
library(knitr)            # Table formatting
library(kableExtra)       # Additional table customization
library(ggpubr)           # Publication-ready plots
library(Amelia)           # Missing data visualization
library(pwr)              # Power analysis
library(DiagrammeR)       # Flowcharts/Diagrams
library(survival)         # Survival analysis
library(survminer)        # Survival plots

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Define the correct file path using the `here` package
data_path <- here("data", "raw-data", "PROMO_Data.csv")

# Read the CSV file, suppressing messages
PROMO_Data <- read_csv(data_path, show_col_types = FALSE)
```

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Data Import & Initial Inspection

# Read the dataset
PROMO_Data <- read_csv(here("data", "raw-data", "PROMO_Data.csv"), show_col_types = FALSE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#  Data Cleaning & Transformation

# Load required libraries (if not already loaded)
library(janitor)   # For clean_names()

# Clean Column Names
PROMO_Data <- PROMO_Data %>% clean_names()

# Remove Empty Rows
PROMO_Data <- PROMO_Data %>%
  filter(!is.na(study_arm), !is.na(stillbirth), !is.na(neonatal_death), !is.na(infant_low_birth_weight_2500_g))

#  Handle Missing Values
PROMO_Data <- PROMO_Data %>%
  replace_na(list(
    study_arm = "Unknown",
    stillbirth = "Unknown",
    neonatal_death = "Unknown",
    placental_malaria = "Unknown",
    delivery_mode = "Unknown"
  )) %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))

#  Convert Data Types
PROMO_Data <- PROMO_Data %>%
  mutate(across(c(study_arm, stillbirth, neonatal_death, placental_malaria, delivery_mode), as.factor)) %>%
  mutate(across(contains("date"), as.Date, format="%Y-%m-%d"))

#  Verify Changes
glimpse(PROMO_Data)  # More readable than str()

#  Summary of Numeric Variables
print(skim(PROMO_Data))  # Explicitly print skim output

#  Frequency Counts of Key Categorical Variables
PROMO_Data %>% count(study_arm, sort = TRUE) %>% print()
PROMO_Data %>% count(placental_malaria, sort = TRUE) %>% print()
PROMO_Data %>% count(preterm_birth_ontoneo, sort = TRUE) %>% print()

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#  Load required libraries
library(dplyr)
library(gtsummary)

#  Select key baseline variables (ensure they exist)
baseline_vars <- c("study_arm", "education_level_omrse", "gravidity", "parity", 
                   "placental_malaria", "preterm_birth_ontoneo", "stillbirth", 
                   "neonatal_death", "infant_low_birth_weight_2500_g")

#  Ensure only existing variables are used
baseline_vars <- intersect(baseline_vars, colnames(PROMO_Data))

# Generate table only if dataset has rows
if (nrow(PROMO_Data) > 0) {
  baseline_characteristics <- PROMO_Data %>%
    select(all_of(baseline_vars)) %>%
    tbl_summary(by = study_arm) %>%
    add_p() %>%
    modify_header(label = "**Variable**") %>%
    bold_labels()

  # Display only the formatted table
  baseline_characteristics
} else {
  cat("️ No data available for analysis.")
}

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load required libraries
library(dplyr)
library(gtsummary)

# Define key variables (check if they exist in dataset)
malaria_outcome_vars <- c("study_arm", 
                           "total_malaria_episodes_during_pregnancy", 
                           "placental_malaria", 
                           "preterm_birth_ontoneo", 
                           "stillbirth", 
                           "infant_low_birth_weight_2500_g")

#  Keep only variables that exist in dataset
malaria_outcome_vars <- intersect(malaria_outcome_vars, colnames(PROMO_Data))

#  Generate the table only if dataset has rows
if (nrow(PROMO_Data) > 0) {
  malaria_outcomes_table <- PROMO_Data %>%
    select(all_of(malaria_outcome_vars)) %>%
    tbl_summary(
      by = study_arm,  # Compare DP vs. SP groups
      statistic = list(
        total_malaria_episodes_during_pregnancy ~ "{mean} ± {sd}",  # Show mean ± SD
        all_categorical() ~ "{n} ({p}%)"  # Show counts and percentages for categorical variables
      ),
      label = list(
        total_malaria_episodes_during_pregnancy ~ "Total Malaria Episodes (Mean ± SD)",
        placental_malaria ~ "Placental Malaria (Histology)",
        preterm_birth_ontoneo ~ "Preterm Birth",
        stillbirth ~ "Stillbirth",
        infant_low_birth_weight_2500_g ~ "Low Birth Weight (<2500g)"
      )
    ) %>%
    add_p() %>%  # Add p-values for comparisons
    modify_header(label = "**Variable**") %>%  # Ensure readable column headers
    bold_labels()  # Bold table labels for clarity

  #  Display only the formatted table
  malaria_outcomes_table
} else {
  cat("️ No data available for analysis.")
}

```
